Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-07 Dail_Mad_ext.tex
intro

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under density-independent exponential population growth, the
model could be written as:
% \begin{equation}
%   \left.
%   \begin{aligned}
%     \mu_t &= N_{t-1}\exp(r) \\ %\label{eq:ss1b}
%     N_t &= \mu_t + \grave{o}_t \\ %\text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
%     X_t &= N_t + \zeta_t %\text{N}(N_t, \zeta_t)
%   \end{aligned} \;
%   \right\} \; \text{for} \; t=2,\hdots,T
%   \label{eq:ss1}
% \end{equation}
\begin{subequations}
  \begin{align}
%    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
    N_1 &= X_1 \label{ss1a} \\
    N_t &= \mu_t + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \quad\;\;\; \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
  \label{eq:ss1}
\end{subequations}
where $\mu_t$ is a deterministic function describing population
growth. For instance, under exponential growth, $\mu_t =
N_{t-1}e^{r}$ where $r$ is the intrinsic rate of
increase. Alternatively, one might consider density-dependent models,
and.  This
flexibility is an appealing feature of state-space models.
Note that no model is
presented for the initial state at $t=1$; we will return to this issue
subsequently. Note also that the exponential
growth model shown in Eq.~\ref{eq:ss1} can be easily replaced with
any population growth model, including density-depedent models. One feature
that is common to most classical state-space models is that the
error terms are assumed to be i.i.d random effects following Gaussian
(Normal) distributions:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

@@

The fact that both abundance ($N_t$) and the observed data ($X_t$) are
typically non-negative integers suggests that it is inappropriate to
model process variation and observation error using Gaussian
distributions. Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

>>>>

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model, $\mu(N_{t-1})$,
coupled with a stochastic model allowing
for random process variation. %, which we denote $\grave{o}_t$.
The deterministic model may be density-dependent, for example
logistic, or it might be density-independent, as in the case of
exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
is the intrinsic rate of increase. The full model can now be written
as:
%As an example, under
%density-independent exponential population growth, the
%model could be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
%    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
%{\flushleft where} $\mu(N_{t-1})$ is the deterministic population
%growth model. %One might alternatively consider density-dependent models
%for $\mu(N_{t-1})$, and this flexibility is an appealing feature of
%state-space models.
In classical state-space models, the two sets of random effects %In most classical state-space models both sets of parameters
are assumed to be i.i.d Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$.
Note that Eq~\ref{eq:ss1a} indicates that no process variation is
present in year 1. This assumption is often made
due to parameter identifiability problems, which we will describe
shortly. First, we turn our attention to the models for the random
effects.

@@

These assumptions are motivated by
convenience rather than biology. Specifically, the Gaussian assumption
allows for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative values of the state-variable, which is
inconsistent with the observed data. The fact that both abundance ($N_t$) and the observed data ($X_t$) are
typically non-negative integers suggests that it is inappropriate to
model process variation and observation error using Gaussian
distributions. Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.
